---
Title of Study: Expanded Case Study on Gun Related Deaths 
Author: Amber P. Giza
File Format and Title: CaseStudy_GunDeaths_2015.Rmd
---

# Step 1: Importing and viewing the data

Similarly to what I did for the past study I conducted back in January of 2025, 
this file and all others will follow the same format. First off, I added the necessary 
imports for the purposes of the R file. After the imports, we can view the csv file 
and data, which shows us 36,252 total rows. 

```{r}
install.packages("ggplot2")
install.packages("tinytex")
install.packages("rmarkdown", dep = TRUE)
install.packages("plotly")
install.packages("DescTools")
install.packages("dplyr")

library(ggplot2)
library(tinytex)
library(rmarkdown)
library(plotly)
library(DescTools)
library(dplyr)

gunDeathsData_2015 <- read.csv("/content/guns_2015.csv", header = TRUE, sep = ",")
View(gunDeathsData_2015)
```

Next, we check to see if there are any nulls inside the data. There were 5 nulls that I found, so 
I had to re-edit the csv file. All 5 NAs/Nulls were in the intent column, so those values were changed
to Undetermined.

```{r}
# Any null data?
sum(is.na(gunDeathsData_2015))
```

Afterwards, I retrieved the next data in the csv files. Since there was only one column
that was an integer/double, I immediately did the analysis for each of the four main measurements:

* Mean
* Median
* Mode
* Standard Deviation

Starting off with the mean, which is close to 43 years old.

```{r}
print("Mean values for each numeric column:")
print("Mean Age: ") # 43.494
mean(gunDeathsData_2015$age)
```

Next is the median, which is less than the mean:

```{r}
# Median Age:
print("Median value for the age column:") # 41
median(gunDeathsData_2015$age)
```

Next is the mode age, which is 24 years old (same value found with Python)

```{r}
# Mode Age:
print("Most occuring value for the age column:") # 24
Mode(gunDeathsData_2015$age)
```

Lastly is Standard Deviation, which can gives us how the age column is statistically varied.

```{r}
# Standard Deviation Age:
print("Standard Deviation of the age:") #19.551 years
SD(gunDeathsData_2015$age)
```

# Graphs and Boxplots

Now come the boxplots! For the five that I generated here:

* African American/Black victims had the most outliers, when measuring their ages
* Hispanic victims had the second most outliers
* Asian/Pacific Islander victims and Native Americans have the same amount of outliers
* White/Caucasian victims had the most "normally" distributed data for age

```{r, echo=FALSE}
# creating boxplot on age and race:
boxplot(gunDeathsData_2015$age ~ gunDeathsData_2015$race, data = gunDeathsData_2015,
        main = "Age of Victims of Gun-related Deaths by Race",
        xlab = "Race",
        ylab = "Age")
```

By looking at the histogram here, we can see that many people around the ages of 20-24
died the most. This area holds the mode as well.

```{r, echo=FALSE}
# Constructing the histogram:

hist(gunDeathsData_2015$age, xlab="Victims of Gun Violence By Age", col = "red", border = "black")
```

I made boxplots here that detailed people's ages along with their gender. Here, both genders of the
victims have no outliers.

```{r, echo=FALSE}
# Creating a boxplot on age and gender:
boxplot(gunDeathsData_2015$age ~ gunDeathsData_2015$sex, data = gunDeathsData_2015,
        main = "Age of Victims of Gun-related Deaths by Gender",
        xlab = "Gender",
        ylab = "Age")
```

This next histogram shows the number of deaths by month. This also shows that July had the most 
victims.

```{r, echo=FALSE}
# Constructing the histogram for number of deaths by month:

hist(gunDeathsData_2015$month, xlab="Total Victims of Gun Violence By Month", col = "blue", border = "black")
```

For the last graph, it's a series of boxplots detailing a person's age and intentions with the gun.
Here:

* Homicides had the most outliers when comapring the ages of the victims
* Accidental deaths, Undetermined Intentions, and Suicides have no outliers
* The Suicide boxplot has the most "normally" distributed data
* Both Accidental and Undetermined Intention boxlots are slightly skewed.

```{r, echo=FALSE}
# Creating a boxplot on age and intentions behind the gun:
boxplot(gunDeathsData_2015$age ~ gunDeathsData_2015$intent, data = gunDeathsData_2015,
    main = "Age of victims of Gun-related Deaths by Intent",
    xlab = "Intentions With the Gun",
    ylab = "Age")
```
# Counting the Non-Numeric data

I had to do the same thing here with counting the percentages of the victims here, like I did with 
the past study. Here, I used dplyr, to count the categorical columns and I'd add find the percentages
by hand. 

```{r}
gunDeathsData_2015 %>%
  count(gunDeathsData_2015$race)
```

## For the percentages: (Total is 36,252 victims)
* Asian/Pacific Islander -> 1.4%
* African American/Black -> 24.66%
* Hispanic -> 9.19%
* Native American -> 0.97%
* White -> 63.78%

Next, I did the same thing for the columns on gender:

```{r}
gunDeathsData_2015 %>%
  count(gunDeathsData_2015$sex)
```

## For the percentages:
* Female -> 14.399%
* Male -> 85.601%

Next, I did the same thing for the columns on intentions with the gun:

```{r}
gunDeathsData_2015 %>%
  count(gunDeathsData_2015$intent)
```

## For the percentages:
* Accidental Deaths -> 1.3%
* Homicides -> 37.12%
* Suicides -> 60.74%
* Undetermined -> 0.84%

Lastly, I calculated the percentages for the months of when the victims died in 2006.

```{r}
gunDeathsData_2015 %>%
  count(gunDeathsData_2015$month)
```

## For the percentages:
* January/1 -> 8.173%
* February/2 -> 6.866%
* March/3 -> 8.38%
* April/4 -> 8.151%
* May/5 -> 8.736%
* June/6 -> 8.372%
* July/7 -> 9.227%
* August/8 -> 8.855%
* September/9 -> 8.429%
* October/10 -> 8.292%
* November/11 -> 8.063%
* December/12 -> 8.456%